Beyond body length: Uncovering the drivers of vertebra number variation in Australian blindsnakes

### 2024-11-08 manuscript.Rmd
update before initial submission

ST: Conceptualisation, Methodology, Formal analysis, Writing - Original draft, reviewing and editing.
AA: Resources, Investigation
SH: Resources, Investigation
SK: Supervision, Funding acquisition, Writing - reviewing and editing.
ES: Conceptualisation, Methodology, Writing - reviewing and editing.

@@

The axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates lack pleomerism, suggesting it is possible to dissociate the number of vertebrae and somatic growth, but the factors that influence this pattern are not well understood. Snakes offer a great model to study axial evolution because they exhibit a wide range of intraspecific variation in vertebra number and interspecific variation in body size, shape and ecology. Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology correlate with vertebra number. To do so, we assembled a comprehensive eco-morphological dataset including X-ray scans for 48 lineages of Australo-Papuan typhlopid snakes. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. We also found that species that have a greater number of vertebrae for their size are found in warmer environments and in harder-to-penetrate soil substrate. These results extend our understanding of the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.

@@

For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope = 0.28, intercept = 256, R^2^ = 0.02, F = 0.74, Z = 0.30, P = 0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which has the greatest maximum TBL, has relatively few vertebrae. We found more phylogenetic signal than expected under Brownnian motion in both maximum number of total vertebrae (Z = 3.70, P = 0.001, K = 1.2; λ = 1.12, log likelihood(λ) = -197.20, LR(λ=0) = 11.82, P = 0.0006) and maximum TBL (Z = 3.02, P = 0.002, K = 1.0; λ = 1.14, log likelihood(λ) = -221.98, LR(λ=0) = 7.42, P = 0.006).

### Relationships between vertebra ratio, number of vertebrae, and aspect ratio
We found that vertebra ratio is positively correlated with mean aspect ratio (PGLS, slope = -0.0048, intercept = 0.73, R^2^ = 0.17, F = 7.09, Z = 2.20, P = 0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-aspect-ratios)). We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope = -1.8, intercept = 120.95, R^2^ = 0.69, F = 79.39, Z = 5.30, P<0.001).

>>>>

ST: Conceptualisation, Methodology, Formal analysis, Writing - Original draft, reviewing and editing.
AA: Resources, Investigation
SH: Resources, Investigation
JSK: Supervision, Funding acquisition, Writing - reviewing and editing.
ES: Conceptualisation, Methodology, Writing - reviewing and editing.

@@

The axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates lack pleomerism, suggesting it is possible to dissociate the number of vertebrae and somatic growth, but the factors that influence this pattern are not well understood. Snakes offer a great model to study axial evolution because they exhibit a wide range of intraspecific variation in vertebra number and interspecific variation in body size, shape and ecology. Here, we investigate the prevalence of pleomerism in a continental radiation of blindsnakes and test whether body shape and ecology correlate with vertebra number. To do so, we assembled a comprehensive eco-morphological dataset including X-ray scans for 48 lineages of Australo-Papuan typhlopid snakes. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. We also found that species that have a greater number of vertebrae for their size are found in warmer environments and in harder-to-penetrate soil substrate. These results extend our understanding of the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.

@@

For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope = 0.28, intercept = 256, R^2^ = 0.02, F = 0.74, Z = 0.30, P = 0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert)). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which has the greatest maximum TBL, has relatively few vertebrae. We found more phylogenetic signal than expected under Brownnian motion in both maximum number of total vertebrae (Z = 3.70, P = 0.001, K = 1.2; λ = 1.12, log likelihood(λ) = -197.20, LR(λ=0) = 11.82, P = 0.0006) and maximum TBL (Z = 3.02, P = 0.002, K = 1.0; λ = 1.14, log likelihood(λ) = -221.98, LR(λ=0) = 7.42, P = 0.006).

### Relationships between vertebra ratio, number of vertebrae, and aspect ratio
We found that vertebra ratio is positively correlated with mean aspect ratio (PGLS, slope = -0.0048, intercept = 0.73, R^2^ = 0.17, F = 7.09, Z = 2.20, P = 0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-aspect-ratios) and Fig. S1). We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope = -1.8, intercept = 120.95, R^2^ = 0.69, F = 79.39, Z = 5.30, P<0.001).
